% Introduction

\chapter{Introduction} \label{sc:intro}

\paragraph{}
Recent developments in machine learning have made significant 
contributions to a wide range of fields that are not 
traditionally considered data science.
Notably the Netflix competition have attracted a collective
effort in developing models that greatly improved prediction
of movie ratings by different users,
creating the best movie recommendation system at the time \cite{FeHeKh12}.
Similar to the Netflix rating data,
student grades in different courses follow the same structure,
allowing the application of the same machine learning techniques.
In this research course, we intend to explore several of the 
machine learning techniques in applications to education.

% \paragraph{}
Specifically, this research project aims to apply machine learning 
to analyze the student grade dataset from \cite{BaRoYo14},
which contains complete transcripts of undergraduate students
from a major Canadian University.
Similar to predicting user ratings,
we are able to predict the grades for courses.
At the same time,
we also will predict student's course and major 
selection choices.
From the predictions, this project intends to analyze
the effect of choosing easier courses on student grades,
specifically by comparing the predicted grades of courses
students did not take against the courses taken 
within the same program.
By analyzing the variation in course difficulty,
these results could potentially improve curriculum design 
for educational institutions and admission procedure for
graduate programs.

This research report will be organized as follow.
Chapter 2 will introduce feed-forward neural networks
in detail, including several techniques that 
can significantly improve training results,
as well as an example case using hand-written digits.
Chapter 3 will introduce two unsupervised learning 
algorithms that can be used to for a different purpose.
Chapter 4 will describe the experiments and 
the results we achieved using the student grade dataset.

% \paragraph{}
% The project will focus on implementing three main methods of inference:
% \vspace{-0em}
% \begin{enumerate}
% \setlength\itemsep{-0.2em}
%   \item Matrix factorization (MF) \cite{FeHeKh12} and if time permits
%   	probabilistic matrix factorization (PMF) \cite{MnSa07, SaMn08}
%   \item Restricted Boltzmann machines (RBM) \cite{Sa09}
%   \item Denoising auto-encoders (DAE) \cite{VLLBM10} and if time permits
%   	variational auto-encoders (VAE) \cite{KiWe13}
% \end{enumerate}


